\myTitle

### 2024-03-18 ckmr-sim-evaluation.Rmd
Got a CKMR figure done.
Still need to get it into the text.

---
title: "Assessing Power of the California Microhap Baseline for GSI (and calculating Fst)"
author: "Eric C. Anderson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true

@@

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    final_baseline = "data/cali-chinook-baseline.rds",
    pop_labels = "inputs/reference-collection-names.csv",
    locus_info = "inputs/ChinookFullPanelLociTable.csv",
    chrom_lengths = "inputs/Otsh_v1.0_chrom_lengths.tsv"
  )
}
```

@@

```{r}
library(tidyverse)
library(CKMRsim)

# get the full baseline with the new markers, etc.  We will toss out some
# non-winter Sacto fish of unknown provenance
full_base <- read_rds(input_list$final_baseline) %>%
  filter(collection != "SacNonW")

@@

# now, we also want to get the genome positions of things.
# we take its position to be the start position of the amplicon
gen_coords <- read_csv(input_list$locus_info) %>%
  select(Locus, AmpliconGenomeV1Location) %>%
  separate(AmpliconGenomeV1Location, into = c("Chrom", "pos_str"), sep = ":") %>%
  separate(pos_str, into = c("Start", "Stop"), sep = "-", convert = TRUE) %>%
  rename(Pos = Start) %>%
  select(-Stop) %>%
  arrange(Chrom, Pos)

# now we have to do some funny stuff to deal with things on unplaced scaffolds.
# We will just sprinkle scaffolds into the existing chromosome lengths from the
# Otsh_v1.0 genome
chrom_lengths <- read_tsv(input_list$chrom_lengths)

# get the unique unplaced scaffolds and assign them randomly to a position
# in an assembled chromosome.  We set it up so that markers on the same scaffold
# will not extend beyond the end of the chromosome
set.seed(5)
rando_coords <- gen_coords %>%
  filter(str_detect(Chrom, "NW_") | is.na(Chrom)) %>%
  group_by(Chrom) %>%
  mutate(max_pos = max(Pos)) %>%
  select(Chrom, max_pos) %>%
  distinct() %>%
  mutate(
    # this gets the chromosome they are assigned to
    ass_chrom = sample(chrom_lengths$chrom, size = n(), prob = chrom_lengths$num_bases, replace = TRUE)
  ) %>% # now we do some stuff to place the scaffold randomly on the chromosome
  left_join(chrom_lengths, by = join_by(ass_chrom == chrom)) %>%
  mutate(
    ass_pos = floor(runif(n = n(), min = 1, max = num_bases - max_pos - 1 ))
  ) %>%
  select(Chrom, ass_chrom, ass_pos) %>%
  ungroup()

# now we have the information to randomly stick the scaffold markers onto
# chromosomes in the genome.  We do this by joining things onto gen_coords
gen_coords2 <- gen_coords %>%
  left_join(rando_coords, by = join_by(Chrom)) %>%
  mutate(
    Pos = case_when(
      is.na(ass_pos) ~ Pos,
      TRUE ~ ass_pos + Pos
    ),
    ass_chrom = case_when(
      is.na(ass_chrom) ~ Chrom,
      TRUE ~ ass_chrom
    )
  ) %>%
  mutate(Chrom = ass_chrom) %>%
  select(Locus, Chrom, Pos)

@@

```{r}
#' @param X a tibble like that nested in a single row of ckmr_sim_nests
#' @param GC2 essentially the gen_coords2 object from above
ckmr_afreqs_ready <- function(X, GC2) {
  tmp <- X %>%
    select(-(repunit:sample_type)) %>%
    pivot_longer(
      -indiv,
      names_to = c("Locus", "gene_copy"),
      values_to = "Allele",
      names_pattern = "^(.*)_([12])$"
    ) %>%
    filter(!is.na(Allele)) %>%
    count(Locus, Allele) %>%
    group_by(Locus) %>%
    filter(n() > 1) %>% # filter out monomorphic loci
    mutate(
      Freq = n / sum(n),
    ) %>%
  ungroup() %>%
  left_join(GC2, by = join_by(Locus)) %>%
  select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq))

@@

# here is one to create a ckmr object.
# Y is simply the output of ckmr_afreqs_ready()
ckmr_create <- function(Y) {
  create_ckmr(
    D = Y,
    kappa_matrix = kappas[c("PO", "FS", "HS", "U"), ],
    ge_mod_assumed = ge_model_TGIE,
    ge_mod_true = ge_model_TGIE,
    ge_mod_assumed_pars_list = list(epsilon = 0.005),
    ge_mod_true_pars_list = list(epsilon = 0.005)
  )
}

# This does the simulate_Qij step and then also estimates the FPRs.
ckmr_simQ <- function(C) {
  simulate_Qij(
    C, 
    calc_relats = c("PO", "FS", "HS", "U"),
    sim_relats = c("PO", "FS", "HS", "U"),
    unlinked = FALSE,
    pedigree_list = pedigrees
  )
}

@@

This is just here while I develop this:
```{r}
X <- ckmr_sim_nests$data[[5]]
GC2 <- gen_coords2

@@

Q <- ckmr_simQ(C)
```

>>>>

---
title: "Assessing Power of the California Microhap Baseline for PBT and relationship inference"
author: "Eric C. Anderson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true

@@

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
    pop_labels = "inputs/reference-collection-names.csv",
    locus_info = "inputs/ChinookFullPanelLociTable_Feb2024Updated.xlsx"
  )
  output_list <- list(
    ckmr_figure = "results/ckmr-sim-eval/fpr-fnr-figure.pdf",
    ckmr_figure_crop = "results/ckmr-sim-eval/fpr-fnr-figure-crop.pdf",
    ckmr_figure_tex = "tex/images/fpr-fnr-figure-crop.pdf"
  )
}

# you can create the necessary output directories like this:
dump <- lapply(output_list, function(x)
  dir.create(dirname(x), recursive = TRUE, showWarnings = FALSE)
)
```

@@

```{r}
library(tidyverse)
library(CKMRsim)
library(readxl)
library(parallel)
library(grid)
library(gridExtra)

# get the full baseline with the new markers, etc.  We will toss out some
# non-winter Sacto fish of unknown provenance
full_base <- read_csv(input_list$final_baseline) %>%
  filter(collection != "SacNonW")

@@

# now, we also want to get the genome positions of things.
# we take its position to be the start position of the amplicon
gen_coords <- read_excel(input_list$locus_info) %>%
  select(Locus, AmpliconGenomveV2Location) %>%
  separate(AmpliconGenomveV2Location, into = c("Chrom", "pos_str"), sep = ":") %>%
  separate(pos_str, into = c("Start", "Stop"), sep = "-", convert = TRUE) %>%
  rename(Pos = Start) %>%
  select(-Stop) %>%
  arrange(Chrom, Pos)

@@

```{r}
#' @param X a tibble like that nested in a single row of ckmr_sim_nests
#' @param GC2 essentially the gen_coords object from above
ckmr_afreqs_ready <- function(X, GC2) {
  tmp <- X %>%
    select(-(repunit:sample_type)) %>%
    pivot_longer(
      -indiv,
      names_to = c("Locus", "gene_copy"),
      values_to = "Allele",
      names_pattern = "^(.*)_([12])$"
    ) %>%
    filter(!is.na(Allele)) %>%
    count(Locus, Allele) %>%
    group_by(Locus) %>%
    filter(n() > 1) %>% # filter out monomorphic loci
    mutate(
      Freq = n / sum(n),
    ) %>%
    ungroup() %>%
    left_join(GC2, by = join_by(Locus)) %>%
    select(Chrom, Pos, Locus, Allele, Freq) %>%
    arrange(Pos, desc(Freq))

@@

# here is one to create a ckmr object.
# Y is simply the output of ckmr_afreqs_ready()
ckmr_create <- function(Y) {
  create_ckmr(
    D = Y,
    kappa_matrix = kappas[c("PO", "FS", "HS", "AN", "U"), ],
    ge_mod_assumed = ge_model_TGIE,
    ge_mod_true = ge_model_TGIE,
    ge_mod_assumed_pars_list = list(epsilon = 0.005),
    ge_mod_true_pars_list = list(epsilon = 0.005)
  )
}

# This does the simulate_Qij step and then also estimates the FPRs.
ckmr_simQ <- function(C, unlinked = FALSE, num_mc_samples = 1e5) {
  simulate_Qij(
    C,
    reps = num_mc_samples,  # 100,000 monte carlo samples by default
    calc_relats = c("PO", "FS", "HS", "AN", "U"),
    sim_relats = c("PO", "FS", "HS", "AN", "U"),
    unlinked = unlinked,
    pedigree_list = pedigrees
  )
}

@@

And here is a function that does the above and then spits out some FPRs and FNRs:
```{r}

get_FPRs_and_FNRs <- function(X, GC2 = gen_coords) {

@@

  # throttle text output to console
  options(CKMRsim.discard_stdout = TRUE)
  options(CKMRsim.linkage_verbosity = 0)

  # Simulate some Q_ij's. Suppress output messages that the CKMRsim package
  # produces
  Q_link <- ckmr_simQ(C) #%>% suppressMessages()
  Q_unl <- ckmr_simQ(C, unlinked = TRUE) #%>% suppressMessages()

  # get FPRs and FNRs for a variety of relationships.  Return in a big tibble

  ret <- list(
    # first for resolving unrelated individuals from PO and FS (and HS, for which there is no hope!)
    all_v_unrel = mc_sample_simple(
      Q = Q_unl,
      Q_for_fnrs = Q_link,
      nu = c("PO", "FS", "HS"),
      de = "U",
      tr = "U",
      method = "IS",
      FNRs = seq(0.01, 0.3, by = 0.01)
    ),
    # then, vanilla MCMC to see how readily we can distinguish FS from PO,
    # and then, of course FS from HS and PO from A, etc.
    po_vs_fs = mc_sample_simple(
      Q = Q_link,
      nu = c("PO"),
      de = "FS",
      tr = "FS",
      method = "vanilla",
      FNRs = seq(0.01, 0.3, by = 0.01)
    ),
    fs_vs_po = mc_sample_simple(
      Q = Q_link,
      nu = c("FS"),
      de = "PO",
      tr = "PO",
      method = "vanilla",
      FNRs = seq(0.01, 0.3, by = 0.01)
    ),
    fs_vs_hs = mc_sample_simple(
      Q = Q_link,
      nu = c("FS"),
      de = "HS",
      tr = "HS",
      method = "vanilla",
      FNRs = seq(0.01, 0.3, by = 0.01)
    ),
    po_vs_hs = mc_sample_simple(
      Q = Q_link,
      nu = c("PO"),
      de = "HS",
      tr = "HS",
      method = "vanilla",
      FNRs = seq(0.01, 0.3, by = 0.01)
    ),
    po_vs_an = mc_sample_simple(
      Q = Q_link,
      nu = c("PO"),
      de = "AN",
      tr = "AN",
      method = "vanilla",
      FNRs = seq(0.01, 0.3, by = 0.01)
    )
  ) %>%
    bind_rows(.id = "what_it_was")

  ret
}

I tried ripping through everything using furrr::future_map() but that did not
actually do it in parallel.  So we are going to just mclapply it:

```{r}
set.seed(5)
tmp_list <- mclapply(
  X = ckmr_sim_nests$data, 
  FUN = function(x) get_FPRs_and_FNRs(x),
  mc.cores = 8
)

ckmr_sim_results <- tibble(
  ckmr_sim_nests,
  fprs = tmp_list
)
```

## Make a plot with all those results

```{r}
# unnest it
ckmr_sim_tib <- ckmr_sim_results %>%
  select(-data) %>%
  unnest(fprs) %>%
  ungroup()

# just print out the different scenarios simulated
ckmr_sim_tib %>%
  distinct(mc_method, numerator, denominator, true_relat)
```

Now, make a figure of those results.

```{r}
# first, try a simple way of categorizing the errors
error_types_tib <- tribble(
  ~"scenario", ~"Error Category",
  "IS-FS-U-U", "Mistaking U for FS",
  "IS-HS-U-U", "Mistaking U for HS",
  "IS-PO-U-U", "Mistaking U for PO",
  "vanilla-FS-HS-HS", "Mistaking HS for FS",
  "vanilla-FS-PO-PO", "Mistaking PO for FS",
  "vanilla-PO-AN-AN", "Mistaking AN for PO",
  "vanilla-PO-FS-FS", "Mistaking FS for PO",
  "vanilla-PO-HS-HS", "Mistaking HS for PO"
)

cst2 <- ckmr_sim_tib %>%
  mutate(scenario = str_c(mc_method, numerator, denominator, true_relat, sep = "-")) %>%
  left_join(error_types_tib, by = join_by(scenario))

source("R/shift_legend.R")

# first do the importance sampling ones
g <- ggplot(cst2 %>% filter(mc_method == "IS"), aes(x = FNR, y = FPR, colour = `Error Category`, linetype = `Error Category`)) +
  geom_line() +
  #geom_segment(aes(x = FNR, xend = FNR, y = FPR - 2 * se, yend = FPR + 2 * se)) +
  facet_wrap(~ ckmr_sim_group, nrow = 4) +
  scale_y_continuous(trans = "log10") +
  theme_bw() + 
  theme(legend.direction = "horizontal") +
  xlab("False Negative Rate") +
  ylab("False Positive Rate (Log10-scaled)") +
  guides(
    colour = guide_legend(
      title.position = "top",
      label.position = "bottom",
      nrow = 2
      )
  ) +
  guides(
    linetype = guide_legend(
      title.position = "top",
      label.position = "bottom",
      nrow = 2
      )
  )

a <- grid.arrange(shift_legend(g))
```

Now, let's look at the vanilla ones:
```{r}
g2 <- ggplot(
  cst2 %>% filter(mc_method == "vanilla" & FPR > 0 & FPR - 2 * sqrt(FPR * (1 - FPR) / 5e4) > 0.8e-05), 
  aes(x = FNR, y = FPR)) +
  geom_segment(
    aes(
      x = FNR, 
      xend = FNR,
      y = FPR - 2 * sqrt(FPR * (1 - FPR) / 5e4),
      yend = FPR + 2 * sqrt(FPR * (1 - FPR) / 5e4),
      colour = `Error Category`
    ),
    linewidth = 0.2,
    show.legend = FALSE
  ) +
  geom_line(aes(colour = `Error Category`, linetype = `Error Category`)) +
  facet_wrap(~ ckmr_sim_group, nrow = 4) +
  scale_y_continuous(trans = "log10") +
  theme_bw() + 
  theme(legend.direction = "horizontal") +
  xlab("False Negative Rate") +
  ylab("False Positive Rate (Log10-scaled)") +
  guides(
    colour = guide_legend(
      title.position = "top",
      label.position = "bottom",
      nrow = 2),
    linetype = guide_legend(
      title.position = "top",
      label.position = "bottom",
      nrow = 2)
  ) +
  theme(legend.direction = "horizontal")

a2 <- grid.arrange(shift_legend(g2))

Then cowplot it:
```{r}
Both <- cowplot::plot_grid(a, a2, nrow = 1, labels = c("a)", "b)"))

ggsave(Both, filename = output_list$ckmr_figure, width = 12, height = 7)

# let's crop that down for use, too
CALL <- paste("pdfcrop ", output_list$ckmr_figure, collapse = " ")
system(CALL)

file.rename(from = output_list$ckmr_figure_crop, to = output_list$ckmr_figure_tex)
````

### 2024-03-18 tex/body-text-of-paper.tex
Got a CKMR figure done.
Still need to get it into the text.

@@

Analysis of the mapping of amplicons for Chr34:828,768 
showed that many ($\approx$89\%) of the reads were off-target (aligning to other chromosomes, 
etc.), making it costly (in terms of sequencing effort) to include the marker in the baseline panel. 
Consequently it was dropped.  The frequency of the different alleles at the two remaining loci
within different reporting units in the baseline are shown in Table~\ref{tab:lfar-freqs}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Allele frequencies across reporting units of the two late-fall associated
markers.  Frequencies are given for the alleles most common amongst late-fall Chinook: nucleotide 
{\tt G} at locus Chr34:865,057 and {\tt A} at Chr34:1,063,084.  $N$ is the total number of gene 
copies available for estimating the allele frequency.}
\label{tab:lfar-freqs}
{\small
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}} lrrrr}
\hline\hline
& \multicolumn{2}{c}{\underline{Chr34:865,057}} & \multicolumn{2}{c}{\underline{Chr34:1,063,084}} \\
Reporting Unit & freq. & $N$ & freq. & $N$ \\ \hline
\input{inputs/lfar-freqs.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule\vspace*{0.3ex}\hrule
\end{table}
%%%%%%%%%%%%%%%
\subsection*{Genetic Variation}

@@

\begin{figure}
\newcommand{\examplecap}{\footnotesize Self-assignment rates  ({\bf a)} and pairwise 
$F_\mathrm{ST}$ values and allele frequencies ({\bf b)} for the collections in the baseline.  The codes of the populations
appear on the outer margins of the table atop cells colored according to the run-timing group
of the collection.  Interior cells in which the column and row correspond to the same reporting
unit are colored according to the reporting unit.  For self-assignment tallies, the row corresponds
to the true source collection and the column corresponds to the assigned collection. In ({\bf b}), for $F_\mathrm{ST}$ values, the upper triangle was computed using the baseline microhaplotype
data. Each cell in the lower triangle shows a small scatter plot of estimated allele frequencies
in the collections.  These frequencies were estimated by, for each locus, randomly
subsampling individuals from each collection to the minimum number of observed alleles in any collection
at the locus.  }
\begin{center}
\includegraphics[width=\columnwidth]{images/gsi_and_fst_fig-crop.pdf}
\end{center}
\caption[\examplecap]{\examplecap}
\label{fig:example}
\end{figure}

>>>>

\subsection*{Population genetic analyses}

@@

Analysis of the mapping of amplicons for Chr34:828,768 
showed that many ($\approx$89\%) of the reads were off-target (aligning to other chromosomes, 
etc.), making it costly (in terms of sequencing effort) to include the marker in the baseline panel. 
Consequently it was dropped.  The frequency of the different alleles at the two remaining loci
within different reporting units in the baseline are shown in Table~\ref{tab:lfar-freqs}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Allele frequencies across reporting units of the two late-fall associated
markers.  Frequencies are given for the alleles most common amongst late-fall Chinook: nucleotide 
{\tt G} at locus Chr34:865,057 and {\tt A} at Chr34:1,063,084.  $N$ is the total number of gene 
copies available for estimating the allele frequency.}
\label{tab:lfar-freqs}
{\small
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}} lrrrr}
\hline\hline
& \multicolumn{2}{c}{\underline{Chr34:865,057}} & \multicolumn{2}{c}{\underline{Chr34:1,063,084}} \\
Reporting Unit & freq. & $N$ & freq. & $N$ \\ \hline
\input{inputs/lfar-freqs.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule\vspace*{0.3ex}\hrule
\end{table}
%%%%%%%%%%%%%%%
\subsection*{Population genetic analyses}

@@

\begin{figure}
\newcommand{\examplecap}{\footnotesize Self-assignment rates  ({\bf a)} and pairwise 
$F_\mathrm{ST}$ values and allele frequencies ({\bf b)} for the collections in the baseline.  The codes of the populations
appear on the outer margins of the table atop cells colored according to the run-timing group
of the collection.  Interior cells in which the column and row correspond to the same reporting
unit are colored according to the reporting unit.  For self-assignment tallies, the row corresponds
to the true source collection and the column corresponds to the assigned collection. In ({\bf b}), for $F_\mathrm{ST}$ values, the upper triangle was computed using the baseline microhaplotype
data. Each cell in the lower triangle shows a small scatter plot of estimated allele frequencies
in the collections---the $x$ value is allele frequencies from the collection in the column and the
$y$ value is from the collection in the row. Frequencies for all alleles are shown, with values within
each cell ranging from 0 to 1.  These frequencies were estimated by, for each locus, randomly
subsampling individuals from each collection to the minimum number of observed alleles in any collection
at the locus.  }
\begin{center}
\includegraphics[width=\columnwidth]{images/gsi_and_fst_fig-crop.pdf}
\end{center}
\caption[\examplecap]{\examplecap}
\label{fig:example}
\end{figure}
